Beyond conceptual eco-physiological hypotheses: insights from a general size and trait-based model of ectotherm thermal performance

### 2018-03-22 draft.Rnw
draft march 22nd

cpG <- ggplot(wtfs) +
  geom_line(aes(x=Temperature, y=length, col=Size,linetype=Scenario),alpha=0.7) +
  theme_cowplot()+
  scale_linetype_discrete(guide='none')+
  scale_color_discrete(guide='none')+
  theme(panel.background = element_blank(),
        plot.background = element_blank(),)+
  labs(x = '',
       y='Relative adult length')

npG <- ggplot(wtfs) +
  geom_line(aes(x=Temperature, y=G/l, col=Size,linetype=Scenario),alpha=0.7) +
  theme(panel.background = element_blank(),
        plot.background = element_blank())+
  labs(x = '',
       y=expression(S~response~y^-1))

>>>>

cpG <- ggplot(wtfs) +
  geom_line(aes(x=Temperature, y=length, linetype=Scenario),alpha=0.7) +
  theme_cowplot()+
  scale_linetype_discrete(guide='none')+
  scale_color_discrete(guide='none')+
  theme(panel.background = element_blank(),
        plot.background = element_blank(),)+
  labs(x = expression("Mean temperature " ( degree*C)),
       y='Relative adult length')

npG <- ggplot(wtfs) +
  geom_line(aes(x=Temperature, y=G/l, col=Size,linetype=Scenario),alpha=0.7) +
  theme(panel.background = element_blank(),
        plot.background = element_blank())+
  labs(x = expression("Mean temperature " ( degree*C)),
       y=expression(S~response~y^-1))

### 2018-03-22 draft_model_only.Rnw
draft march 22nd

There is currently no accepted universal concept or general mechanistic model that explains thermal performance and ecological responses to temperature (\cite{lefevre_are_2016, jutfelt_oxygen-and_2018}). Two main theories are debated as potential general therories describing temeprature influences on ectotherms. The theory of Oxygen and Capacity Limitation of Temperature (OCLTT; \cite{portner_oxygen-and_2010}),  based on the general idea of oxygen limitation for major ecological processes (growth, fitness, etc; see below), provides a conceptual framework rather than a mechanistic theoretical framework, and its major assumptions have been called into question on numerous occasions (\cite[e.g., ][]{lefevre_are_2016, jutfelt_oxygen-and_2018}). The gill-oxygen limitation theory (GOLT) aims to provide a physiological justification for the use of a temperature and oxygen scaled version of the von Bertalanffi growth model as a mechanistic model of thermal performance. The validity of gill limitation of oxygen supply, as well as the ultimate limitation of growth by oxygen in all conditions, is strongly debated (\cite{brander_overconfidence_2013,pauly_sound_2017,lefevre_modelling_2017,lefevre_models_2017}). In addition, the von Bertalanffi growth model only encapsulates a highly abstract view of organism bio-energetics and ecology, with no reference to ecological rates such as food aquisition, mortality and reproduction. As such, it can only represent a very limited set of aspects of species ecology, and cannot on its own account for changes in ecological rates that are often observed with temperature (\cite{englund_temperature_2011,rall_universal_2012}).

@@

where $f(w,T)$ is the feeding level ([0,1]) at weight $w$ and temperature $T$, determined by consumption rate $\gamma w^p \Theta$ (search rate $\gamma w^p$ times prey availability $\Theta$), maximum consumption $h w^q$, and $\tau$, which can be interpreted as the proportion of time spent foraging (or proportion of maximum attack rate) or a multiplyer on a basal foraging speed, such that at $\tau=1$, foraging occurs at maximum speed. In the following, we will refer to $\tau$ as the activity fraction for sake of generality. Maximum consumption, determined by digestive (enzymatic) processes (\cite{jeschke_predator_2002, sentis_parsing_2013}), is assumed to scale with temperature according to the Arrhenius scaling $c_T = e^{E_a(T-T_0)/k \times T \times T_0}$. Available energy $P(w,T)$ is determined by supply ($S(w,T)$) from prey consumption ($f(w,T) h c_T w^q$), with $\beta$ a loss due to specific dynamic action (SDA, or heat increment; the energy spent absorbing food), and $\phi$ is the fraction of food excreted and egested. Metabolic demands ($D(w,T)$) are those of standard metabolism ($k w^n$), as well as active metabolism, scaled in units of standard metabolism as $\delta k w$, with $\delta$ the factorial aerobic scope (FAS). The activity fraction $\tau$ determines the proportion of total metabolic costs that are due to active metabolism. Note that we assume that active metabolism scales directly with $w$, owing to muscular demands scaling approximately isometrically with weight (\cite{glazier_activity_2009,brett1965relation}). In practice, the scaling may not be strictly isometric owing to reduced drag at larger size (\cite[e.g., ][]{ware_bioenergetics_1978}), as well as non-strictly isometric growth of muscle mass with weight. However, the precise exponents of standard and foraging metabolism only alter predictions quantitatively, but do not change the over-all results.

@@

\hline
\addlinespace
\multicolumn{4}{c}{\textbf{Biomass Metabolism}} \\
Activity coefficient & $\delta$ & \multicolumn{2}{c}{4}    \\
Specific dynamic action & $\beta$ & \multicolumn{2}{c}{0.15} \\
Egestion and excretion & $\phi$ & \multicolumn{2}{c}{0.25}  \\
Coeff. for std.~metabolism  & $k$ ($g\cdot y^{-1}$)  & 1 &  1.5 \\
Exponent for std.~metabolism & $n$ & 0.88 & 0.75  \\

\addlinespace
\multicolumn{4}{c}{\textbf{Feeding ecology}}\\
Coeff. for Encountered food & $\gamma\Theta$ ($y^{-1}$)& \multicolumn{2}{c}{60 (40/80)} \\
Exponent for clearance rate $\gamma$ & $p$ & \multicolumn{2}{c}{0.8}  \\
Coeff. for Maximum consumption rate & $h$ ($y^{-1}$) & 30 & 60  \\
Exponent for max.~consumption $\gamma$ & $q$ & \multicolumn{2}{c}{0.8} \\
Coeff. for constant mortality & $M$ ($g\cdot y^{-1}$) & 0.1  & 1  \\
Coeff. for activity related mortality & $\rho$ ($y^{-1}$) & 6 & 1\\
Exponent for mortality & $\nu$ & \multicolumn{2}{c}{0.2} \\

@@

mstar_ss <- plot_data_growth_tO2(ss)$mstar
mstar_fs <- plot_data_growth_tO2(fs)$mstar

@@

gouta <- parallel::mclapply(scsa,plot_data_growth_tO2,mc.cores = 3)
goutr <- parallel::mclapply(scs,plot_data_growth_tO2,mc.cores = 2)
@

@@

Our quantitative thermal impact model generalises existing eco-physiological models for particular species and stocks (\cite{hufnagl_physiological_2011, holt_climate_2014,holt_climate_2015}), and allows to develop a more nuanced understanding of interactions between temperature, oxygen limitation and ecology for species with varying traits. In line with the conceptual framework of Fry's aerobic scope and the OCLTT, our model suggests that oxygen limitation can be a potentially important ecological driver, especially at extreme temperatures for species with declining MOSs in such temperature regimes. At the onset of this limitation, ecological parameters change drastically, and growth, mortality are strongly impacted. This limitation closely mimics limitations seen in wild fish (\cite{myrick_temperature_2000}), and is in line with observations that fish often seek specific water temperatures to optimise metabolic function (\cite{armstrong_diel_2013,claireaux_physiology_1995}). Fitness, however, appears to be limited through the metabolic scope primarily via limitations at temperature extremes and impact on particular life-history stages. For instance, oxygen limitation is a more severe constraint for small individuals (Figure \ref{fig:size-rates}), and thereby can limit growth performance early in life, impacting over-all fitness.

@@

Resource availibility imposes a strong environmental constraint on organisms, with all aspects from optimal activity levels, mortality and available energy for growth ultimately influenced by available prey. Changes in prey availibility thus influece the temperature response directly via available energy, and indirectly, through altered energetic requirements to procure food and changes in mortality due to changes in the optimal acitivity level. In low-food environments, changing energetic demands with temeprature are not easily adjusted for, and any adjustment demands higher energic costs and mortality risk. This environment-driven change in costs of temperature adjustments leads to the strong modulation of the growth repsonse, as well as changing gradients in fitness with temperature in these environments. In particular, in low-food environments, asymptitic size is strongly reduced at high temperatures, whereas this is not necessarily the case in high-food environments. Increased growth and mortality changes at low temperatures lead to a more strongly domed fitness curve.

@@

The maximum oxygen consumption is the oxygen consumption during maximal activity level that can be sustained over some time, and is usually termed the maximum metabolic rate (MMR). Although the MMR is often used synonymously with both maximum oxygen supply and demand, in some species the maximal oxygen consumption ($D^max_{O_2}(T)$) is not reached at maximum activity levels, but rather during digestion (\cite{priede_metabolic_1985}). We use MMR and oxygen supply interchangeably, although maximum oxygen demand (i.e., at full activity) may be lower than the MMR. We assume that oxygen supply is temperature dependent and we follows a flexible dome-shaped function (\cite{gnauck2013freshwater, lefrancois_influence_2003}):

\begin{align}
f_{O_2}(T)&=\lambda(T)\left(1-e^{(O_2(T)-P_{crit})\log(0.5)/(P50-P_{crit}})\right(,\\
\lambda(T)&=\zeta(\frac{T_{\text{max}}-T}{T_{\text{max}}-T_{\text{opt}}})^\eta \times \exp(-\eta\frac{T_{\text{max}}-T}{T_{\text{max}}-T_{\text{opt}}}),
\end{align}

>>>>

There is currently no accepted universal concept or general mechanistic model that explains ectotherm thermal performance and ecological responses to temperature (\cite{lefevre_are_2016, jutfelt_oxygen-and_2018}). In fish, two main theories are debated as potential general therories describing temeprature influences on ectotherms. The theory of Oxygen and Capacity Limitation of Temperature (OCLTT; \cite{portner_oxygen-and_2010}),  based on the general idea of oxygen limitation for major ecological processes (growth, fitness, etc; see below), provides a conceptual framework rather than a mechanistic theoretical framework, and its major assumptions have been called into question on numerous occasions (\cite[e.g., ][]{lefevre_are_2016, jutfelt_oxygen-and_2018}). The gill-oxygen limitation theory (GOLT) aims to provide a physiological justification for the use of a temperature and oxygen scaled version of the von Bertalanffi growth model as a mechanistic model of thermal performance. The validity of gill limitation of oxygen supply, as well as the ultimate limitation of growth by oxygen in all conditions, is strongly debated (\cite{brander_overconfidence_2013,pauly_sound_2017,lefevre_modelling_2017,lefevre_models_2017}). In addition, the von Bertalanffi growth model only encapsulates a highly abstract view of organism bio-energetics and ecology, with no reference to ecological rates such as food aquisition, mortality and reproduction. As such, it can only represent a very limited set of aspects of species ecology, and cannot on its own account for changes in ecological rates that are often observed with temperature (\cite{englund_temperature_2011,rall_universal_2012}).

@@

where $f(w,T)$ is the feeding level ([0,1]) at weight $w$ and temperature $T$, determined by consumption rate $\gamma w^p \Theta$ (search rate $\gamma w^p$ times prey availability $\Theta$), maximum consumption $h w^q$, and $\tau$, which can be interpreted as the proportion of time spent foraging (or proportion of maximum attack rate) or a multiplyer on a basal foraging speed, such that at $\tau=1$, foraging occurs at maximum speed. In the following, we will refer to $\tau$ as the activity fraction for sake of generality. Maximum consumption, determined by digestive (enzymatic) processes (\cite{jeschke_predator_2002, sentis_parsing_2013}), is assumed to scale with temperature according to the Arrhenius scaling $c_T = e^{E_a(T-T_0)/k \times T \times T_0}$. Available energy $P(w,T)$ is determined by supply ($S(w,T)$) from prey consumption ($f(w,T) h c_T w^q$), with $\beta$ a loss due to specific dynamic action (SDA, or heat increment; the energy spent absorbing food), and $\phi$ is the fraction of food excreted and egested. Metabolic demands ($D(w,T)$) are those of standard metabolism ($k w^n$), as well as active metabolism, scaled in units of standard metabolism as $\delta k w$, with $\delta$ determining the maginitude of activity cost relative to standard metabolic costs. The activity fraction $\tau$ determines the proportion of total metabolic costs that are due to active metabolism. Note that we assume that active metabolism scales directly with $w$, owing to muscular demands scaling approximately isometrically with weight (\cite{glazier_activity_2009,brett1965relation}). In practice, the scaling may not be strictly isometric owing to reduced drag at larger size (\cite[e.g., ][]{ware_bioenergetics_1978}), as well as non-strictly isometric growth of muscle mass and other relevant factors with weight (\cite{killen_little_2007}). However, the precise exponents of standard and foraging metabolism only alter predictions quantitatively, but do not change the over-all results.

@@

\hline
\addlinespace
\multicolumn{4}{c}{\textbf{Biomass Metabolism}} \\
Activity coefficient & $\delta$ ($g^{n}$) & \multicolumn{2}{c}{4}    \\
Specific dynamic action & $\beta$ & \multicolumn{2}{c}{0.15} \\
Egestion and excretion & $\phi$ & \multicolumn{2}{c}{0.25}  \\
Coeff. for std.~metabolism  & $k$ ($g^{1-n}\cdot y^{-1}$)  & 1 &  1.5 \\
Exponent for std.~metabolism & $n$ & 0.88 & 0.75  \\

\addlinespace
\multicolumn{4}{c}{\textbf{Feeding ecology}}\\
Coeff. for Encountered food & $\gamma\Theta$ ($g^{1-p} \cdot y^{-1}$)& \multicolumn{2}{c}{60 (40/80)} \\
Exponent for clearance rate $\gamma$ & $p$ & \multicolumn{2}{c}{0.8}  \\
Coeff. for Maximum consumption rate & $h$ ($g^{1-q}\cdot y^{-1}$) & 30 & 60  \\
Exponent for max.~consumption $h$ & $q$ & \multicolumn{2}{c}{0.8} \\
Coeff. for constant mortality & $M$ ($g\cdot y^{-1}$) & 0.1  & 1  \\
Coeff. for activity related mortality & $\rho$ ($y^{-1}$) & 6 & 1\\
Exponent for mortality & $\nu$ & \multicolumn{2}{c}{0.2} \\

@@

#mstar_ss <- plot_data_growth_tO2(ss)$mstar
#mstar_fs <- plot_data_growth_tO2(fs)$mstar

mstar_ss <- 50
mstar_fs <- 50

@@

gouta <- parallel::mclapply(scsa,plot_data_growth_tO2,mstar=50,mc.cores = 3)
goutr <- parallel::mclapply(scs,plot_data_growth_tO2,mstar=50,mc.cores = 2)
@

@@

Our quantitative thermal impact model generalises existing eco-physiological models for particular species and stocks (\cite{hufnagl_physiological_2011, holt_climate_2014,holt_climate_2015}), and allows to develop a more nuanced understanding of interactions between temperature, oxygen limitation and ecology for species with varying traits. In line with the conceptual framework of Fry's aerobic scope and the OCLTT, our model suggests that oxygen limitation can be a potentially important ecological driver, especially at extreme temperatures for species with declining MOSs in such temperature regimes. At the onset of this limitation, ecological parameters change drastically, and growth, mortality are strongly impacted. This limitation closely mimics limitations seen in wild fish (\cite{myrick_temperature_2000}), and is in line with observations that fish often seek specific water temperatures to optimise metabolic function (\cite{armstrong_diel_2013,claireaux_physiology_1995}). Fitness, however, appears to be limited through the metabolic scope primarily via limitations at temperature extremes and impact on particular life-history stages. For instance, oxygen limitation is a more severe constraint for small individuals (Figure \ref{fig:size-rates}), and thereby can limit growth performance early in life, impacting over-all fitness. Also, changes in environmental oxygen supply, if beyond an oganisms ability to compensate via passive or active compensation mechanisms, will induce an over-all lower aerobic scope and lead to an earlier onset of oxygen limitation, but such scenarios do not change the qualitative predictions from our model.

@@

Resource availibility imposes a strong environmental constraint on organisms, with all aspects from optimal activity levels, mortality and available energy for growth ultimately influenced by available prey. Changes in prey availibility thus influence individual temperature response directly via available energy, and indirectly, through altered energetic requirements to procure food and changes in mortality due to changes in the optimal acitivity level. In low-food environments, changing energetic demands with temperature are not easily adjusted for, and any adjustment demands higher energic costs and mortality risk. This environment-driven change in costs of temperature adjustments leads to the strong modulation of the growth repsonse, as well as changing gradients in fitness with temperature in these environments. In particular, in low-food environments, asymptotic size is strongly reduced at high temperatures, whereas this is not necessarily the case in high-food environments. Increased growth and mortality changes in such environments then combine to produce a more strongly domed fitness curve than in environemts where food supply is not limiting.

@@

The maximum oxygen consumption is the oxygen consumption during maximal activity level that can be sustained over some time, and is usually termed the maximum metabolic rate (MMR). Although the MMR is often used synonymously with both maximum oxygen supply and demand, in some species the maximal oxygen consumption ($D^{max}_{O_2}(T)$) is not reached at maximum activity levels, but rather during digestion (\cite{priede_metabolic_1985}). We use MMR and oxygen supply interchangeably, although maximum oxygen demand (i.e., at full activity) may be lower than the MMR. We assume that oxygen supply is temperature dependent and follows a flexible dome-shaped function (\cite{gnauck2013freshwater, lefrancois_influence_2003}):

\begin{align}
f_{O_2}(T)&=\lambda(T)\left(1-e^{(O_2(T)-P_{crit})\log(0.5)/(P50-P_{crit})}\right),\\
\lambda(T)&=\zeta\left(\frac{T_{\text{max}}-T}{T_{\text{max}}-T_{\text{opt}}}\right)^\eta \times \exp\left(-\eta\frac{T_{\text{max}}-T}{T_{\text{max}}-T_{\text{opt}}}\right),
\end{align}
